---
output: html_document
date: "2023-10-06"
name: "Ryan Dotzlaw"

---


### Ryan Dotzlaw

#### Q1
##### Consider the system of equations Ax=b, where:

```{r}
A = matrix(c(c(2,1,3,0,5),c(3,0,-2,3,0),c(2,-27,6,2,1)), nr=5, nc=3)
A
b = c(-1,2,0,1,-2)
b
```
##### Find a QR factorization of A

A QR factorization of 'A' consists of two matrices, 'Q' and 'R', where 'Q' is an orthonormal matrix consisting of the vectors generated from the Gram-Schmidt process, and where 'R' is an upper triangular matrix, with each entry 'x_ij' being the dot product between the vector 'e_i' from the Gram-Schmidt process, and the vector 'a_j' from the matrix 'A'


Firstly, use the Gram-Schmidt process on 'A'

```{r}
# function for finding the norm of a vector (default function 'norm' was giving errors)
normal = function(vec){
  return(sqrt(sum(vec^2)))
}

# function for projecting 'x' onto 'v'
project = function(x, v){
  dot = x %*% v
  nsqr = normal(v)^2
  res = (dot / nsqr)
  res = res * v
  return(res)
}

# the matrix 'Q', same size as 'A', right now filled w/ 'NA'
Q = matrix(nr=nrow(A), nc=ncol(A))

# v1 is equal to the first column in 'A'
v1 = A[,1]
# v2 is equal to the second column in 'A', minus the projection of that column onto v1
v2 = A[,2] - project(A[,2], v1)
# v3 is equal to the third column in 'A', minus the projection of that column onto v1, and minus the projection of that column onto v2
v3 = A[,3] - project(A[,3], v1) - project(A[,3], v2)

# now normalize the vectors, and place into 'Q'
Q[,1] = v1 / normal(v1)

Q[,2] = v2 / normal(v2)
Q[,3] = v3 / normal(v3)

Q

# from there, R = Q^T A, so
R = round(t(Q) %*% A, 15)
R

# and you can get A from multiplying Q and R
round(Q %*% R,15)

# using the qr method...
QR = qr(A)
# gives the same result
round(qr.Q(QR) %*% qr.R(QR), 15)


```
I would like to note, that above I rounded the calculation of R and A to the 15th digit due to the fact that the zero entries in R and A were instead +/-1e-16 instead of zero. But the fact that all the non-zero entries were correct led me to believe that it was just an error in the decimal arithmetic, so I just rounded the results.

##### Now find the least squares solution

The least squares solution using QR factorization is: x~ = R^-1 Q^T b

```{r}

# using my 'Q' and 'R'
xtilde1 = solve(R) %*% t(Q) %*% b
xtilde1

A %*% xtilde1

# using the 'qr' method
xtilde2 = solve(qr.R(QR)) %*% t(qr.Q(QR)) %*% b
xtilde2
A %*% xtilde2

```


#### Q2

##### Find the best approximation to a solution to the following system:
* x + y - z = 2
* -y + 2z = 6
* 3x +2y - z = 11
* -x + z = 0

To find the best approximation, we need to find the 'x~' such that 'Ax~' is as close to 'b' as possible.

This is pretty standard at this point, treat it as a least squares problem and solve for 'x~' using: x~ = inv(t(A)A)t(A)b

```{r}
# the system of equations can be represented by a matrix and a vector:

# where each column in 'A' represents the coefficients of the x, y, and z variables
A = matrix(c(c(1,0,3,-1), c(1,-1,2,0), c(-1,2,-1,1)), nc=3, nr=4)
A

# where 'b' is the values each equation is equal to
b = c(2,6,11,0)
b

# solve for x~
xtilde = solve(t(A) %*% A) %*% t(A) %*% b
xtilde

# see how close the solution is, if it wasn't an approximation this should give a zero vector
A %*% xtilde - b


```



#### Q5 (GS computation only)

##### Apply the Gram-Schmidt process to the basis of the subspace 'W' to obtain an orthonormal basis of 'W'

The subspace 'W' is described as the span of the following vectors:

```{r}
x1 = c(1,2,-2,1)
x1
x2 = c(1,1,0,2)
x2
x3 = c(1,8,1,0)
x3
```

Now to calculate the orthonormal basis of 'W'

```{r}
# u1 is just x1 normalized
u1 = x1 / normal(x1)

# v2 is x2 minus the projection of x2 onto x1
v2 = x2 - project(x2, x1)
# normalize, and round b/c there's a floating point error again
u2 = round(v2 / normal(v2),15)

# v3 is x3 minus the projection of x3 onto x1 and the projection of x3 onto x2
v3 = x3 - project(x3, x1) - project(x3, x2)
# normalize
u3 = v3 / normal(v3)

# so the orthonormal basis of 'W' is the following three vectors:
u1
u2
u3

```


#### Q6

##### Find an orthonormal basis for Real^3 that contains a vector parallel to:

```{r}
v1 = c(1,2,3)
```

Since the orthonormal basis must contain a vector parallel to 'v1', this means that the vector in the orthonormal basis is a scalar multiple of 'v1', logically if follows that this will be the normalized version of 'v1'

```{r}
u1 = v1 / normal(v1)
u1
```

Next we need to find any other vector that is orthogonal to 'v1', such as:

```{r}
v2 = c(1,1,-1)
v1 %*% v2
```
Since the dot product is zero, we know that these two vectors are orthogonal, additionally the projection of 'v2' onto 'v1' will be the zero vector due to this

```{r}
project(v2,v1)
```
So if we're applying the Gram-Schmidt process to 'v2', it will just result in 'v2'

Now we can normalize 'v2'

```{r}
u2 = v2 / normal(v2)
```

Now we apply to Gram-Schmidt process to any vector to get one that is orthogonal to both 'v1' and 'v2'

```{r}
x3 = c(1,0,0)
v3 = x3 - project(x3, v1) - project(x3, v2)
v3

# then normalize
u3 = v3 / normal(v3)
```
Now we have an orthonormal basis for Real^3, that contains a vector parallel to 'v1' as follows:

```{r}
u1
u2
u3
```



