---
output: html_document
date: "2023-11-28"
name: "Ryan Dotzlaw"
---
#### Ryan Dotzlaw

Set up

```{r}
if (!require("markovchain")) {
    install.packages("markovchain")
    library(markovchain)
}
if (!require("igraph")) {
    install.packages("igraph")
    library(igraph)
}
```


#### Q1

```{r}
# set up helper functions

# limit of 'p^k' k->inf (actually, 1e5), is p^k positive
reg = function(p){
  pk = p
  for(i in c(1, 1e5)){
    if(reg1(pk)){
      # compute stationary probability
      # normalize eigen vector associated w/ eigen value '1' (always e$vectors[,1])
      e = eigen(p)
      # normalize eigen vector
      norm_e = abs(e$vectors[,1]) / sum(abs(e$vectors[,1]))
      # return(c(norm_e, e$vectors[,j]))
      return(norm_e)
      
    }
    pk = p %*% p
  }
  # didn't reach limit in reasonable time period
  return(FALSE)
}

# if matrix positive, return true
reg1 = function(p_in){
  p = p_in %*% p_in
  # loop thru rows
  for(i in c(1, length(p[1,]))){
    # loop thru cols
    for(j in c(1, length(p[,1]))){
      # entry non-positve
      if(p[i,j] <= 0){
        # entire matrix non-positive -> non-regular markov-chain
        return(FALSE)
      }
    }
  }
  
  return(TRUE)
}

# a markov chain has absorbing states if:
# some value on the main diagonal is '1' -> stuck at that state
# returns false if 'p' is not absorbing,
# returns the absorbing states otherwise
absorb = function(p, l=TRUE){
  # assuming p is valid (cols sum up to 1)
  val = FALSE
  count = 1
  for(i in c(1:length(p[,1]))){
    if(p[i,i] == 1){
      # absorbing state: 'i'
      if(typeof(val) == "logical"){ # first absorbing state
        val = c()
      }
      if(l){
        val[count] = paste("s",toString(i),sep="")
      } else {
        val[count] = i
      }
      count = count + 1
    }
  }
  return(val)
}

```



```{r}
# Define matrices
p1 = matrix(c(c(1,0),c(0,1)), nrow=2, ncol=2)
p2 = matrix(c(c(0.2, 0.5, 0.3), c(0, 0.5, 0.5), c(0.5, 0.5, 0)), nrow=3, ncol=3)
p3 = matrix(c(c(0.1, 0.9, 0, 0), c(0.8, 0.1, 0.1, 0), c(0, 0.1, 0.3, 0.6), c(0, 0, 0.5, 0.5)), nrow=4, ncol=4)
p1
p2
p3
```

Plot the graphs.

```{r}
G1 = make_graph(c(c(1,1), c(2,2)))
E(G1)$weights = c(1,1)
par(mar=c(0,0,0,0)+.1)
plot(G1)

G2 = make_graph(c(
  c(1,1), c(1,2),
  c(1,3), c(2,2), c(2,3),
  c(3,1), c(3,2)))
E(G2)$weights = c(0.2, 0.5, 0.3, 0.5, 0.5, 0.5, 0.5)
par(mar=c(0,0,0,0)+.1)
plot(G2, edge.arrow.size=E(G2)$weights*10, layout=layout_in_circle(G2))

G3 = make_graph(c(
  c(1,1), c(2,2), c(3,3), c(4,4),
  c(1,2), c(2,1), c(2,3), c(3,2),
  c(3,4), c(4,3)))
E(G3)$weights = c(0.1, 0.1, 0.3, 0.5, 0.9, 0.8, 0.1, 0.1, 0.6, 0.5)
par(mar=c(0,0,0,0)+.1)
plot(G3, edge.arrow.size=E(G3)$weights*10, layout=layout_in_circle(G3))
```


```{r}
# is p1 regular?
reg(p1)
# is p2 regular?
reg(p2)
# is p3 regular?
reg(p3)

```

So p1 is not regular, but p2 and p3 are, and their stationary distribution vectors are the normalized eigen vectors associated with the eigen value '1'.

```{r}

# is p1 absorbing?
absorb(p1)
# is p2 absorbing?
absorb(p2)
# is p3 absorbing?
absorb(p3)

```

So only 'p1' is absorbing, with the absorbing states: {1, 2}

#### Verify Q1 with 'markovchain' Library
```{r}
m1 = new('markovchain',
         transitionMatrix=p1,
         states=c('s1','s2'), byrow=FALSE)
m1
is.regular(m1)
absorbingStates(m1)
# so P1 is not regular, but is absorbing, with the absorbing states {1, 2}, so my Q1 is right

m2 = new('markovchain',
         transitionMatrix=p2,
         states=c('s1','s2', 's3'), byrow=FALSE)
m2
is.regular(m2)
steadyStates(m2)
absorbingStates(m2)
# so P2 is regular w/ same stationary probability, and isn't absorbing, so my Q1 is right

m3 = new('markovchain',
         transitionMatrix=p3,
         states=c('s1','s2', 's3', 's4'), byrow=FALSE)
m3
is.regular(m3)
steadyStates(m3)
absorbingStates(m3)
# so P3 is regular w/ same stationary probability, and isn't absorbing, so my Q1 is right
```

So all my Q1 answers are correct.

#### Q2

Create standard form function

```{r}
# converts markov chain matrix to standard form
sForm = function(p){
  
  states = absorb(p, FALSE) # list of absorbing states
  if(typeof(states) == "logical"){ # no absorbing states
    return(FALSE)
  }
  
  
  # col swap
  # move the 'states' columns into the front
  M1 = p[,states]

  # now add non-absorbing columns
  M1 = cbind(M1, p[,-states])
  
  # print(M1)
  
  # row swap
  # move the '1's in the absorbing state columns to the top to create an identity matrix
  M2 = M1[states,] # grab row from M1

  # now add non-absorbing rows
  M2 = rbind(M2, M1[-states,])
  
  return(M2)
}

val_comp = function(sf, a){
  sf_l = length(sf[1,])
  x = sf_l - length(a)
  # print(x)
  # print(sf_l)
  I = diag(nrow=x ,ncol=x)
  Q = sf[(x+1):sf_l, (x+1):sf_l]
  
  N = solve(I-Q) # (I-Q)^-1
  
  Ti = list()
  for(r in c(1:length(N[,1]))){ # for each row
    Ti[r] = sum(N[r,])
  }
  
  R = sf[1:length(a), (x+1):sf_l] # gets sf: rows [1,length(a)), cols [length(a), length(sf[,1]))
  
  B = R %*% N
  
  print(R)
  print(N)
  
  # print('---- N is ---- ')
  # print(N)
  # print('---- Ti is ---- ')
  # print(Ti)
  # print('---- B is ---- ')
  # print(B)
  
  return(c(as.data.frame(N), as.data.frame(Ti), as.data.frame(B)))
}
```


```{r}
P = matrix(
  c(0.2, 0.3, 0.4, 0.1, 0, 1, 0, 0, 0.5, 0.2, 0.2, 0.1, 0, 0, 0, 1), 
  nrow=4, ncol=4)
P

a = absorb(P)
a
```

```{r}
sf = sForm(P)
sf
```

```{r}
vc = val_comp(sf, a)
# N
as.data.frame(vc[1:2])
# Ti
as.data.frame(vc[3:4])
# B
as.data.frame(vc[5:6])


```

```{r}
M = new("markovchain",
        transitionMatrix=P,
        states=c('s1','s2','s3','s4'), byrow=FALSE)
M
canonicForm(M)

```

Which is the same as my standard form matrix, so my Q2 computes the right standard form matrix.

