---
output: html_document
date: "2023-10-12"
name: "Ryan Dotzlaw"
---

#### Ryan Dotzlaw

#### Q1
##### Compute the SVD of the following matrices

```{r}
A = matrix(c(c(1,0), c(1,0)), nc=2, nr=2)
A
B = matrix(c(c(-2,0), c(0, 0)), nc=2, nr=2)
B
C = matrix(c(c(2,0), c(0,2), c(1,0)), nc=3, nr=2)
C
```
##### The SVD formula
The SVD of any matrix 'M', size 'm' by 'n' is:

M = U * E * V^T,

where U is the orthonormal eigenvectors of M * M^T,

and where V is the orthonormal eigenvectors of M^T * M,

and where E is a diagonal matrix of the singular values (square root of eigenvalues) of size 'm' by 'n'

##### SVD of A

```{r}
# calculating U
AAt = A %*% t(A)
eu = eigen(AAt)
# for some reason the eigenvectors are negative, I assume this has something to do with the eigen() implementation
# it should be noted that below, when using svd(), U has no negative values.
# So, since U needs to be the orthonormal eigenvectors of 'AAt', we can just multiply both eigenvectors
# by -1, and they will remain orthonormal eigenvectors of 'AAt' associated with the same eigenvalues

Ua = matrix(c(-1 * eu$vectors), nc=2, nr=2)
Ua

# calculating V
AtA = t(A) %*% A
ev = eigen(AtA)
Va = matrix(c(ev$vectors), nc=2, nr=2)
Va

# calculate E
# looking at the eigenvalues
eu$values
ev$values
# the only non-zero eigenvalue is 2,
# so the only singular value put in the E matrix is sqrt(2)
# so E is just sqrt(2) on the main diagonal, and the rest is zeros
# E needs to be size 'm' by 'n', so 2 by 2
Ea = matrix(c(c(sqrt(eu$values)), c(0, 0)), nc=2, nr=2)
Ea

# now the result of U %*% E %*% t(V) should be equal to A again
A == Ua %*% Ea %*% t(Va)
```
```{r}
# using the SVD function built in to calculate svd of A
svdA = svd(A)
# U is
svdA$u
# V is
svdA$v
# the singular values of A are
svdA$d
# set up the diagonal E matrix using the non-zero singular values svdA$d
# E is 'm' by 'n', so 2 by 2, with the singular values on the diagonal
E = diag(svdA$d, nc=2, nr=2)
E
res = svdA$u %*% E %*% t(svdA$v)
res
A

# on my machine, this returns FALSE when comparing the 1's in 'res' to the 1's in 'A'
# I assume this is some floating point arithmetic error, but the value are both 1's of 'num' type
# I legitimately don't know why it does this, but taking the ceiling of both matrices works,
# so it must be some floating point error?
res == A
# what????
ceiling(res) == ceiling(A)
```
##### SVD of B
```{r}
# calculate U
BBt = B %*% t(B)
eu = eigen(BBt)
# once again, the eigenvectors are negative for some reason, in this case we want a negative to get the '-2' in B
Ub = matrix(c(eu$vectors), nc=2, nr=2)
Ub

# calculate V
BtB = t(B) %*% B
ev = eigen(BtB)
# these eigenvectors are all negative too, 
# but since they're already negative in U, this will cause the 'SVD = -B'
# so multiply these eigenvectors by -1 (they're still othonormal eigenvectors of BtB)
Vb = matrix(c(-1 * ev$vectors), nc=2, nr=2)
Vb

# calculate E
eu$values
ev$values
Eb = diag(c(c(sqrt(eu$values))), nc=2, nr=2)
Eb

res = Ub %*% Eb %*% t(Vb)
res
res == B
```
```{r}
# using built in SVD to calculate svd of B
svdB = svd(B)
# U is
svdB$u
# V is
svdB$v
# singular values of B are
svdB$d

# now create the diagonal E matrix using the singular values on the main diagonal
# E has the same dimensions as B, so 2 by 2
E = diag(svdB$d, nc=2, nr=2)
E

resB = svdB$u %*% E %*% t(svdB$v)
resB
B
resB == B

```
##### SVD of C
```{r}
# calculate U
CCt = C %*% t(C)
eu = eigen(CCt)
# negative again, for some reason, you know the drill by now
Uc = matrix(c(-1 * eu$vectors), nc=2, nr=2)
Uc

# calculate V
CtC = t(C) %*% C
ev = eigen(CtC)
# the second eigenvector is negative for some reason
ev$vectors[,2]
# make it non-negative, to make the SVD work (remains orthonormal eigenvector)
ev$vectors[,2] = -1 * ev$vectors[,2]
Vc = matrix(c(ev$vectors), nc=3, nr=3)
Vc

# calculate E
eu$values
# floating point error, round eigenvalues
round(ev$values, 14)
# non-zero eigenvalues are 5 and 4
# size of E is the same as C, so a 2 x 3 matrix
Ec = diag(sqrt(eu$values), nc=3, nr=2)
Ec

resC = Uc %*% Ec %*% t(Vc)
resC
C
# weird floating point stuff??
round(resC) == round(C)
```

```{r}
# using built in SVD to calculate svd of C
svdC = svd(C)
# U is
svdC$u
# V is
svdC$v

# this only gives two eigenvectors for V
# this is because the eigenvalue associated with the third eigenvector is 0
# so the third eigenvector is technically arbitrary, but I included it for matrix size consistency


# the singular values of C are
svdC$d

# now create E, diagonal matrix w/ same dimensions as C, so a 2 x 3 matrix
# however, since 'svdC$v' only contains 2 vectors, E will have to be a 2 x 2 matrix
E = diag(svdC$d, nc=2, nr=2)
E

resC = svdC$u %*% E %*% t(svdC$v)
resC
C

# float stuff again
round(resC) == round(C)
```
#### Q2
##### Compute the Pseudo-Inverses of the matrices given in Q1

The pseudo-inverse of a matrix M uses the SVD of that matrix.

Given the SVD of M: M = U * E * V^T,

The pseudo-inverse of M, 'M^t' is:

M^t = V * E^t * U^T

Where E^t is the pseudo inverse of E, specifically, a diagonal matrix of the reciprocal of the non-zero singular values of 'M'

##### Pseudo-Inverse of A
```{r}
# have U and V, need to make E^t

# get the eigenvalues
sing = eigen(A %*% t(A))$values
# invert the non-zero entries, take the square root
sing = sqrt(c(sing[sing != 0]^-1, sing[sing == 0]))
sing
# make diagonal matrix of non-zero singular reciprocals
Et = diag(sing, nc=2, nr=2)
Et

At = Va %*% Et %*% t(Ua)
At

```

##### Pseudo-Inverse of B

```{r}
# have U and V, need to make E^t

# get the eigenvalues
sing = eigen(B %*% t(B))$values
# invert the non-zero entries, take the square root
sing = sqrt(c(sing[sing != 0]^-1, sing[sing == 0]))
sing
# make diagonal matrix of non-zero singular reciprocals
Et = diag(sing, nc=2, nr=2)
Et

Bt = Vb %*% Et %*% t(Ub)
Bt
```


##### Pseudo-Inverse of C

```{r}
# have U and V, need to make E^t

# get the eigenvalues
sing = eigen(C %*% t(C))$values
# invert the non-zero entries, take the square root
sing = sqrt(c(sing[sing != 0]^-1, sing[sing == 0]))
sing
# make diagonal matrix of non-zero singular reciprocals
# needs to be 3 x 2 to matrix multiply w/ V
Et = diag(sing, nc=2, nr=3)
Et

Ct = Vc %*% Et %*% t(Uc)
Ct
```

